Peer Assessment 1
========================================================

Load the required packages
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

Load the data and convert the date variable:
```{r loaddata}
dat <- read.csv("activity.csv")
dat$date <- as.Date(dat$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r totalsteps, message=FALSE}
totals <- summarise(group_by(dat, date), totalsteps=sum(steps, na.rm=TRUE))
head(totals)
```

2. Make a histogram of the total number of steps taken each day
```{r totalstepshist, fig.height=4}
qplot(date, data=totals, weight=totalsteps, geom="histogram", binwidth=1)
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean_median}
tsmean <- round(mean(totals$totalsteps), 2)
tsmedian <- median(totals$totalsteps)
```
The mean of the total number of steps taken per day is `r tsmean`, and the median is `r tsmedian`.

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r averagesteps, fig.height=4}
averages <- summarise(group_by(dat, interval), average_steps=mean(steps, na.rm=TRUE))
p <- ggplot(averages, aes(x=interval, y=average_steps))
p + geom_line() + scale_x_continuous(breaks=seq(0,2250,250))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r maxinterval}
maxinter <- averages$interval[which.max(averages$average_steps)]
```
The 5-interval that contains the maximum number of steps, on average, is `r maxinter`.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r numberofNA}
numNA <- sum(is.na(dat$steps))
```
The total number of missing values is `r numNA`.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated.   
The following is a function that replaces a NA value with the average of steps for the interval:
```{r fillNAfunction}
fillNA <- function(dat){
  if (is.na(dat$steps)){
    dat$steps <- averages$average_steps[averages$interval == dat$interval]
    return(dat)
  }
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r newdataset, cache=TRUE}
newdat <- dat
repind <- which(is.na(newdat$steps))
for (i in repind){
  newdat[i,] <- fillNA(newdat[i,])
}
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r newtotalsteps, fig.height=4}
newtotals <- summarise(group_by(newdat, date), totalsteps=sum(steps))
qplot(date, data=newtotals, weight=totalsteps, geom="histogram", binwidth=1)
newmean <- round(mean(newtotals$totalsteps), 2)
newmedian <- median(newtotals$totalsteps)
```
The new value of the mean of the total number of steps taken per day is `r newmean`, and the new value of the median is `r newmedian`.

Do these values differ from the estimates from the first part of the assignment? 
```{r differ}
meandiff <- abs(tsmean-newmean)
mediandiff <- abs(tsmedian-newmedian)
```
The difference between the first mean and the new one is `r meandiff` steps, and the difference between the first median and the new one is `r mediandiff`.

What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r difftotal}
totaldiff <- sum(newtotals$totalsteps - totals$totalsteps)
```
When imputing missing data, the total number of steps increases in `r totaldiff` steps.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r weekdays}
enddays <- c("Saturday", "Sunday")
dayvec <- weekdays(newdat$date)
newdat$day[dayvec %in% enddays] <- "weekend"
newdat$day[!(dayvec %in% enddays)] <- "weekday"
newdat$day <- as.factor(newdat$day)
head(newdat)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r plotweekdays, fig.height=4}
newaverages <- summarise(group_by(newdat, interval, day), average_steps=mean(steps, na.rm=TRUE))
p <- ggplot(newaverages, aes(x=interval, y=average_steps))
p + geom_line() + scale_x_continuous(breaks=seq(0,2250,250)) + facet_grid(day ~ .)
```
